---
title: "ScDiagnostics BioC2024 Workshop"
author: Anthony Christidis^[anthony-alexander_christidis@hms.harvard.edu]
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ScDiagnostics BioC2024 Workshop}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::knit_hooks$set(pngquant = knitr::hook_pngquant)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "ragg_png",
  dpi = 72,
  fig.retina = 2,
  fig.width = 8.6667,
  fig.height = 5.3334,
  fig.align = "center",
  out.width = "100%",
  pngquant = "--speed=1 --quality=1-10"
)
```

# ScDiagnostics BioC2024 Workshop

Authors: Anthony Christidis[^1], Andrew Ghazi[^2], Ludwig
Geistlinger[^3]. <br/> Last modified: 22 July, 2024

[^1]: Core for Computational Biomedicine, Harvard Medical School

[^2]: Core for Computational Biomedicine, Harvard Medical School

[^3]: Core for Computational Biomedicine, Harvard Medical School

## Overview

### Description

`scDiagnostics` is a comprehensive toolkit designed for the analysis and
diagnostics of single-cell RNA sequencing (scRNA-seq) data. This package
provides functionalities for comparing principal components, visualizing
canonical correlation analysis (CCA) outputs, and plotting cell
type-specific MDS and PCA projections, among other tasks.

### Pre-requisites

Before attending this workshop, participants should have a basic
understanding of single-cell RNA sequencing (scRNA-seq) data analysis
and familiarity with Bioconductor packages. Specifically, knowledge of
the `SingleCellExperiment` package will be useful for working with
single-cell datasets. While familiarity with dimensionality reduction
techniques, such as Principal Component Analysis (PCA), and statistical
tests will enhance the learning experience, it is not required.

A solid foundation in single-cell analysis techniques and tools will
significantly benefit those looking to fully leverage the capabilities
of specialized packages like `scDiagnostics`." Basic proficiency in R
and R Markdown is also recommended to follow along with the workshop
exercises.

## Setup (*R* / *Bioconductor*)

First, let's load the necessary libraries and data.

```{r libraries, message=FALSE}
# Load necessary libraries
library(scDiagnostics)
library(scran)
library(scater)
```

We will use the following datasets available in `scDiagnostics` for the
workshop. These datasets are processed versions of the dataset
`HeOrganAtlasData(tissue = c("Marrow"))` in the Bioconductor package
`scRNAseq`.

```{r data, message=FALSE}
# Load data
data("reference_data")
data("query_data")
```

We will also create a dataset where we extract only one cell type (CD4)
from the data. We use the SingleR annotation to subset the data. THis
will allow us to evaluate how well aligned the CD4 cells are between the
reference and query datasets.

```{r subset, message=FALSE}
# Subset data for CD4
reference_data_subset <- reference_data[, which(reference_data$expert_annotation == "CD4")]
query_data_subset <- query_data[, which(query_data$SingleR_annotation == "CD4")]

# Selecting highly variable genes
ref_top_genes <- getTopHVGs(reference_data_subset, n = 500)
query_top_genes <- getTopHVGs(query_data_subset, n = 500)

# Intersect the gene symbols to obtain common genes
common_genes <- intersect(ref_top_genes, query_top_genes)
reference_data_subset <- reference_data_subset[common_genes,]
query_data_subset <- query_data_subset[common_genes,]

# Run PCA
reference_data_subset <- runPCA(reference_data_subset)
query_data_subset <- runPCA(query_data_subset)
```

### Time outline

An example for a 45-minute workshop:

| Activity                               | Time |
|----------------------------------------|------|
| Introduction to Annotation Diagnostics | 10m  |
| Brief Overview                         | 10m  |
| Example Workflow                       | 15m  |
| Questions and Answers                  | 10m  |

### Learning goals

The goal of this workshop is to provide participants with a thorough
understanding of the `scDiagnostics` package, enabling them to
effectively analyze and interpret single-cell RNA sequencing data.
Participants will gain hands-on experience in utilizing the package's
diverse functionalities to enhance their single-cell data analysis
workflows.

By the end of this workshop, participants will understand the
capabilities and applications of `scDiagnostics`, master visualization
techniques for both multiple and single cell types, calculate
correlations and distances between cell types, perform and interpret
statistical tests, detect and visualize anomalies, evaluate data quality
and gene overlap, and utilize additional analytical tools.

### Learning objectives

By the end of this workshop, participants will be able to:

1.  **Visualize Differences Between Datasets:**
    -   Use various plotting functions to compare query and reference
        datasets across multiple and single cell types.
2.  **Evaluate Gene Overlap:**
    -   Utilize tools for gene overlap measures.
3.  **Calculate Correlations and Distances:**
    -   Implement functions to calculate correlations and distances
        between different cell types.
4.  **Perform Statistical Tests:**
    -   Apply statistical methods to compare groups and assess
        significance.
5.  **Detect Anomalies:**
    -   Identify and visualize anomalies in single-cell data.

## Brief Overview of scDiagnostics

### Visualize Differences Between Datasets

#### `plotCellTypePCA`

This function plots the principal components for different cell types in
the query and reference datasets.

```{r pca_plot, message=FALSE}
pc_plot <- plotCellTypePCA(query_data = query_data, 
                           reference_data = reference_data,
                           cell_types = c("CD4", "CD8", "B_and_plasma", "Myeloid"),
                           query_cell_type_col = "SingleR_annotation", 
                           ref_cell_type_col = "expert_annotation", 
                           pc_subset = 1:5)
pc_plot
```

#### `boxplotPCA`

This function generates a boxplot visualization of principal components
(PCs) for different cell types across two datasets (query and
reference).

```{r boxplot, message=FALSE}
pc_boxplot <- boxplotPCA(query_data = query_data, 
                         reference_data = reference_data,
                         query_cell_type_col = "SingleR_annotation", 
                         ref_cell_type_col = "expert_annotation", 
                         cell_types = c("CD4", "CD8", "B_and_plasma", "Myeloid"),
                         pc_subset = 1:5)
pc_boxplot
```

#### `plotCellTypeMDS`

This function facilitates the assessment of similarity between reference
and query datasets through Multidimensional Scaling (MDS) scatter plots.
It allows the visualization of cell types based on a dissimilarity
matrix computed from a user-selected gene set.

```{r mds, message=FALSE}
mds_plot <- plotCellTypeMDS(query_data = query_data, 
                            reference_data = reference_data, 
                            cell_types = c("CD4", "CD8", "B_and_plasma"),
                            query_cell_type_col = "SingleR_annotation", 
                            ref_cell_type_col = "expert_annotation")
mds_plot
```

#### `comparePCA`

This function compares the principal components (PCs) obtained from
separate PCA on reference and query datasets for a single cell type
using either cosine similarity or correlation.

The S3 plot method generates a heatmap to visualize the cosine
similarities between principal components from the output of the
comparePCA function.

```{r pca_comparison, message=FALSE}
similarity_mat <- comparePCA(query_data = query_data_subset, 
                             reference_data = reference_data_subset, 
                             query_cell_type_col = "SingleR_annotation", 
                             ref_cell_type_col = "expert_annotation",
                             pc_subset = 1:5)
plot(similarity_mat)
```

### Evaluate Gene Overlap

#### `calculateVarImpOverlap`

This function uses the Random Forest algorithm to calculate the
importance of genes in differentiating between cell types within both a
reference dataset and a query dataset. The function then compares the
top genes identified in both datasets to determine the overlap in their
importance scores.

```{r var_imp_overlap, message=FALSE}
var_imp <- calculateVarImpOverlap(reference_data = reference_data, 
                                    query_data = query_data, 
                                    ref_cell_type_col = "expert_annotation", 
                                    query_cell_type_col = "SingleR_annotation")
var_imp[["var_imp_comparison"]]
```

### Calculate Correlations and Distances

#### `calculateAveragePairwiseCorrelation`

Computes the average pairwise correlations between specified cell types
in single-cell gene expression data.

The S3 plot method takes the output of the
`calculateAveragePairwiseCorrelation` function, which should be a matrix
of pairwise correlations, and plots it as a heatmap.

```{r pairwise_correlations, message=FALSE}
cor_matrix_avg <- calculateAveragePairwiseCorrelation(query_data = query_data, 
                                                      reference_data = reference_data, 
                                                      query_cell_type_col = "SingleR_annotation", 
                                                      ref_cell_type_col = "expert_annotation", 
                                                      cell_types = c("CD4", "CD8", "B_and_plasma"), 
                                                      pc_subset = 1:5,
                                                      correlation_method = "spearman")
plot(cor_matrix_avg)
```

### Perform Statistical Tests

#### `calculateNearestNeighborProbabilities`

This function computes the probabilities for each query cell of
belonging to either the reference or query dataset for each cell type
using nearest neighbor analysis.

The S3 plot method generates a density plot showing the distribution of
probabilities for each cell of belonging to either the reference or
query dataset for each cell type.

```{r probabilities, message=FALSE}
nn_output <- calculateNearestNeighborProbabilities(query_data = query_data, 
                                                   reference_data = reference_data,
                                                   query_cell_type_col = "SingleR_annotation", 
                                                   ref_cell_type_col = "expert_annotation",
                                                   pc_subset = 1:5)
plot(nn_output)
```

#### `calculateHotellingPValue`

This function calculates Hotelling's T-squared statistic for comparing
multivariate means between reference and query datasets, projected onto
a subset of principal components (PCs). It performs a permutation test
to obtain p-values for each cell type specified.

```{r hotelling, message=FALSE}
p_values <- calculateHotellingPValue(query_data = query_data, 
                                     reference_data = reference_data, 
                                     query_cell_type_col = "SingleR_annotation", 
                                     ref_cell_type_col = "expert_annotation",
                                     pc_subset = 1:5) 
round(p_values, 5)
```

### Detect Anomalies

#### `detectAnomaly`

This function projects the query data onto the PCA space of the
reference data. An isolation forest is then built on the reference data
to identify anomalies in the query data based on their PCA projections.
If no query dataset is provided by the user, the anomaly scores are
computed on the reference data itself. Anomaly scores for the data with
all combined cell types are also provided as part of the output.

The S3 plot method extracts the specified PCs from the given anomaly
detection object and generates scatter plots for each pair of PCs. It
uses ggplot2 to create a faceted plot where each facet represents a pair
of PCs. Anomalies are highlighted in red, while normal points are shown
in black.

```{r anomaly, message=FALSE}
anomaly_output <- detectAnomaly(reference_data = reference_data, 
                                query_data = query_data, 
                                ref_cell_type_col = "expert_annotation", 
                                query_cell_type_col = "SingleR_annotation",
                                pc_subset = 1:5,
                                n_tree = 500,
                                anomaly_treshold = 0.6) 
plot(anomaly_output, 
     cell_type = "CD4", 
     pc_subset = 1:5, 
     data_type = c("query", "reference"))
```

## Example Workflow

First let us see how well the `SingleR` annotation performed

```{r expert_comparison, message=FALSE}
table(Expert_Annotation = query_data$expert_annotation, 
      SingleR = query_data$SingleR_annotation)

```

We can see that according to expert annotation, a lot of CD8 cells were
incorrectly labeled as CD4 cells by `SingleR`. Let's identify some of
these cells (the top 5 deviating cells for CD4).

First, we compute the distances from each query cell to each reference
cell.

```{r distances, message=FALSE}
distance_data <- calculateCellDistances(query_data = query_data, 
                                        reference_data = reference_data, 
                                        query_cell_type_col = "SingleR_annotation", 
                                        ref_cell_type_col = "expert_annotation",
                                        pc_subset = 1:10) 
```

Now, let's compute the top six anomalies for CD4 using the anomaly
detection function.

```{r anomaly_cd4, message=FALSE}
cd4_anomalies <- detectAnomaly(reference_data = reference_data, 
                               query_data = query_data, 
                               query_cell_type_col = "SingleR_annotation", 
                               ref_cell_type_col = "expert_annotation",
                               pc_subset = 1:10,
                               n_tree = 500,
                               anomaly_treshold = 0.5)$CD4
cd4_top6_anomalies <- names(sort(cd4_anomalies$query_anomaly_scores, decreasing = TRUE)[1:6])
```

Next, we plot the densities of the distances for each of these anomalous
cells for CD4 and CD8.

```{r plot_anomaly_cd4, message=FALSE}
plot(distance_data, ref_cell_type = "CD4", cell_names = cd4_top6_anomalies)
plot(distance_data, ref_cell_type = "CD8", cell_names = cd4_top6_anomalies)
```

We can also compute some measures of similarity between these cells and
all cell types.

```{r anomaly_overlap, message=FALSE}
overlap_measures <- calculateCellDistancesSimilarity(query_data = query_data, 
                                                     reference_data = reference_data, 
                                                     cell_names = cd4_top6_anomalies,
                                                     query_cell_type_col = "SingleR_annotation", 
                                                     ref_cell_type_col = "expert_annotation",
                                                     pc_subset = 1:10) 
overlap_measures
```
